Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2023-11-04 docs/manuscript/paper.tex
major clean up c.f. reproducibility
initial fixes

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling. The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods. It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered and tested for accuracy. There is a bias of $\sim 0.3$ rad in the marginalised one-dimensional posterior for the orbital inclination $\iota$, introduced by dropping the so-called `pulsar terms'. Smaller biases $\lesssim 1 \%$ are also observed in other static parameters.
\end{abstract}

@@

\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
%\begin{figure*}
%	%\centering % Not needed
%	\begin{subfigure}[b]{0.49\textwidth}
%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
%		\label{fig:6MB_BFS}
%	\end{subfigure}  
%    \hfill
%	\begin{subfigure}[b]{0.49\textwidth}
%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
%		\label{fig:25MB_bfs}
%	\end{subfigure}
%	\caption{
%		}
%	\label{fig:four figures}
%\end{figure*}

@@

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.

@@

\section{Representative end-to-end analysis} \label{sec:rep_example}
In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}. To assist with reproducibility, we present intermediate outputs from each step. In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$. In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter. In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline

@@

By not setting priors on $\gamma^{(n)}$ and $d^{(n)}$ we reduce the number of parameters to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f. Equation \eqref{eq:psrparams}. Explicitly, we write
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ . \label{eq:psrreduced}
\end{equation}
The injected static parameters and their corresponding priors for the representative analysis in Section \ref{sec:rep_example} are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected value & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
		\hline
		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		\vspace{1mm} &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
		\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column). The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f. Equation \eqref{eq:psrreduced}.
}
	\label{tab:parameters_and_priors}
\end{table*}

\subsection{Earth and pulsar terms}\label{sec:earth_psr_terms}
The general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}, separates into two cosine terms. The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars. The argument of the cosine corresponds to the GW phase at the Earth.  The second term, $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d \right \}$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and varies between pulsars. The argument of the cosine corresponds to the GW phase at each individual pulsar. The first and second terms are commonly referred to as the ``Earth term'' and ``pulsar term'' respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the inferred sky position \citep{Zhupulsarterms,Chen2022}. In this work we follow the standard approach and drop the pulsar terms from the inference calculation (but not the source model). Explicitly the measurement equation used in the Kalman filter reduces to
\begin{equation}
		f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
		\label{eq:measuremen_earth}
	\end{equation}
	with
	\begin{equation}
		g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
		\label{eq:g_func_trig_earth}
	\end{equation}
	We defer the inclusion of the pulsar terms in the inference calculation to a future paper. As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem because the measurement equation, Equation \eqref{eq:g_func_trig_earth}, is no longer a function of the pulsar distance.  We stress that the pulsar terms are dropped only when doing Bayesian inference, i.e. from the Kalman filter model that feeds into the nested sampling algorithm. The synthetic data include the pulsar term in full. \newline

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. Hence it is evident for this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, that the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ . \newline

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$ is generally smaller than the prior domain.}
	\label{tab:Wasserstein}
\end{table}

>>>>

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single gravitational wave source in the pulse times of arrival. A two-step inference procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling. The procedure estimates the static parameters in the problem, such as the sky position of the source, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods. It also returns the Bayes factor relating a model with a single gravitational wave source to one without, complementing traditional detection methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered and tested for accuracy. There is a bias of $\sim 0.3$ rad in the marginalised one-dimensional posterior for the orbital inclination $\iota$, introduced by dropping the so-called `pulsar terms'. Smaller biases $\lesssim 1 \%$ are also observed in other static parameters.
\end{abstract}

@@

\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}

@@

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ s$^{-3/2}$ for PSR J1939$+$2134.

@@

\section{Representative end-to-end analysis} \label{sec:rep_example}
In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}. To assist with reproducibility, we present intermediate outputs from each step. In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$. In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter. In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline

@@

By not setting priors on $\gamma^{(n)}$ and $d^{(n)}$ we reduce the number of parameters to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f. Equation \eqref{eq:psrparams}. Explicitly, we write
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ . \label{eq:psrreduced}
\end{equation}
The injected static parameters and their corresponding priors for the representative analysis in Section \ref{sec:rep_example} are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected value & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
		\hline
%		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
		 & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		\vspace{1mm} $\boldsymbol{\theta}_{\rm psr}$ &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
		\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column). The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f. Equation \eqref{eq:psrreduced}.
}
	\label{tab:parameters_and_priors}
\end{table*}

\subsection{Earth and pulsar terms}\label{sec:earth_psr_terms}
The general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}, separates into two cosine terms. The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars. The argument of the cosine corresponds to the GW phase at the Earth.  The second term, $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d \right \}$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and varies between pulsars. The argument of the cosine corresponds to the GW phase at each individual pulsar. The first and second terms are commonly referred to as the ``Earth term'' and ``pulsar term'' respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the inferred sky position \citep{Zhupulsarterms,Chen2022}. In this work we follow the standard approach and drop the pulsar terms from the inference calculation (but not the source model). Explicitly the measurement equation used in the Kalman filter reduces to
\begin{equation}
		f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
		\label{eq:measuremen_earth}
	\end{equation}
	with
	\begin{equation}
		g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
		\label{eq:g_func_trig_earth}
	\end{equation}
	We defer the inclusion of the pulsar terms in the inference calculation to a future paper. As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem because the measurement equation, Equation \eqref{eq:g_func_trig_earth}, is no longer a function of the pulsar distance.  We stress that the pulsar terms are dropped only when doing Bayesian inference, i.e. from the Kalman filter model that feeds into the nested sampling algorithm. The synthetic data include the pulsar terms in full. \newline

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ . \newline

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$ is generally smaller than the prior domain.}
	\label{tab:Wasserstein}
\end{table}
